---
title: "Get started: a prior-focused Bayesian workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started: a prior-focused Bayesian workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%")
options(mc.cores = 2)
```

bmbeR walks a regression model through the stages of a Bayesian workflow
(Gelman et al., 2020), with the prior treated as a first-class part of the
model rather than an afterthought:

1. **Specify** priors on a meaningful scale.
2. **Check** what they imply *before* seeing the outcome (prior predictive check).
3. **Fit** the model and **verify** the computation (modern MCMC diagnostics).
4. **Quantify** how much the prior influenced the result (power-scaling sensitivity).
5. **Evaluate** predictions with proper scoring rules.
6. **Report** everything in a checklist aligned with the Bayesian Analysis
   Reporting Guidelines (Kruschke, 2021).

Models are fitted with [rstanarm](https://mc-stan.org/rstanarm/), so every
result is an ordinary `stanreg` object that works with the rest of the Stan
ecosystem.

```{r, message = FALSE}
library(bmbeR)
data(kidiq, package = "rstanarm")
head(kidiq)
```

The `kidiq` data (Gelman & Hill, 2007) record the test scores of 434 children
together with their mothers' IQ and whether the mother finished high school.
We hold out a test set to evaluate predictions at the end.

```{r}
set.seed(42)
test_rows <- sample(nrow(kidiq), 100)
train <- kidiq[-test_rows, ]
test  <- kidiq[test_rows, ]
```

## 1. Specify priors

Priors only make sense relative to the units of the data. Children's test
scores are on an IQ-like scale (mean about 100, SD about 15), so:

* the **intercept** is the expected score of a child whose mother has average
  IQ and schooling (rstanarm puts the intercept prior at the predictor means,
  so you don't have to centre the predictors yourself): `normal(100, 20)`;
* a one-point increase in the mother's IQ is unlikely to change the child's
  score by more than about two points: `normal(0, 1)` for `mom_iq`;
* finishing high school might plausibly shift scores by up to about 20 points:
  `normal(0, 10)` for `mom_hs`;
* the residual SD is probably of the order of 15 points: `exponential(1/15)`.

```{r}
priors <- prior_config(
  intercept = prior_spec("normal", location = 100, scale = 20),
  slope     = prior_spec("normal", location = 0, scale = c(1, 10)),
  aux       = prior_spec("exponential", rate = 1 / 15)
)
priors
```

If you have no substantive knowledge to encode, leave components as `NULL`
to use rstanarm's autoscaled weakly informative defaults, or derive priors
from data with `empirical_bayes_priors()` (see the article on
*Data-informed priors*).

## 2. Check the priors before fitting

A prior predictive check simulates data from the priors alone. Test scores
cannot be negative, and scores above 200 would be absurd, so we declare
that range as plausible:

```{r}
pc <- prior_predictive_check(train, kid_score ~ mom_iq + mom_hs,
                             prior_config = priors,
                             plausible_range = c(0, 200), refresh = 0)
pc
plot(pc)
```

The simulated data sets are wider than the observed data, as they should be
(the priors are weakly informative), but almost all simulated values stay
within the plausible range.

## 3. Fit the model and verify the computation

```{r}
fit <- fit_model_with_prior(train, kid_score ~ mom_iq + mom_hs,
                            prior_config = priors,
                            chains = 4, iter = 1000, refresh = 0)
attr(fit, "bmb_convergence")
```

`fit_model_with_prior()` runs `check_convergence()` automatically using the
criteria of Vehtari et al. (2021): rank-normalised R-hat below 1.01, bulk-
and tail-ESS of at least 100 per chain, no divergent transitions and
adequate E-BFMI. If a check fails it warns and still returns the fit so you
can investigate; rank plots are the most sensitive visual check:

```{r}
plot(attr(fit, "bmb_convergence"), type = "rank")
```

## 4. How much did the prior matter?

`plot_prior_posterior()` overlays each prior on its posterior:

```{r}
plot_prior_posterior(fit)
```

`prior_sensitivity()` quantifies the same question without refitting, by
power-scaling the prior and the likelihood (Kallioinen et al., 2023):

```{r}
sens <- prior_sensitivity(fit)
sens
```

```{r, include = FALSE}
flagged <- sens$summary$variable[sens$summary$diagnosis != "-"]
conflict <- sens$summary$variable[sens$summary$diagnosis == "prior-data conflict"]
```

How to read the table:

* `contraction` near 1 means the data have sharpened the prior a lot.
* The `*_sens` columns measure how much the posterior mean and SD respond to
  strengthening the prior or the likelihood (0 = not at all).
* `conflict_z` measures disagreement between prior and data in standard-error
  units; values beyond ±2 indicate a prior-data conflict.

`r if (length(flagged) == 0) "No parameter is sensitive to the prior: the likelihood dominates, so the conclusions do not hinge on the exact priors chosen." else paste0("The diagnosis flags ", paste0("\x60", flagged, "\x60", collapse = " and "), ": for ", if (length(flagged) == 1) "this parameter" else "these parameters", " the prior contributes a non-negligible share of the posterior precision. ", if (length(conflict) == 0) "Because every |conflict_z| is well below 2, prior and data agree, which is what an informative-but-sensible prior should look like. The remaining parameters are dominated by the likelihood." else "For the parameters labelled as a conflict, prior and data disagree and the prior deserves a second look.")`
The *Prior sensitivity* article shows what weakly informative, conflicting
and weakly identified cases look like.

## 5. Evaluate predictions

```{r}
perf <- evaluate_model_performance(fit, test)
perf
```

Besides RMSE and MAE of the posterior mean, bmbeR reports proper scoring
rules that also reward honest uncertainty: the held-out log predictive
density (ELPD) and the continuous ranked probability score (CRPS). Coverage
of the 90% predictive intervals close to 90% indicates well-calibrated
uncertainty.

## 6. Report

```{r}
report <- workflow_report(fit, prior_check = pc, sensitivity = sens,
                          performance = perf)
report
```

Items marked `[check]` need attention and `[ -- ]` items have not been done
yet. The report adds posterior predictive p-values and PSIS-LOO
cross-validation, which are cheap to compute from the fit.

## Where next

* *Choosing priors*: why scale matters, autoscaling, and prior predictive
  checks for logistic regression.
* *Data-informed priors*: unit-information, empirical Bayes and power priors,
  and why naively re-using the data is a mistake.
* *Convergence diagnostics*: what each criterion means and what failure looks like.
* *Prior sensitivity*: the theory behind the diagnostics, and refitting under
  alternative priors.
* *Predictive evaluation*: proper scoring rules and model comparison on held-out data.

## References

Gelman, A., & Hill, J. (2007). *Data Analysis Using Regression and
Multilevel/Hierarchical Models*. Cambridge University Press.

Gelman, A., Vehtari, A., Simpson, D., et al. (2020). Bayesian workflow.
*arXiv:2011.01808*.

Kallioinen, N., Paananen, T., Bürkner, P.-C., & Vehtari, A. (2023). Detecting
and diagnosing prior and likelihood sensitivity with power-scaling.
*Statistics and Computing*, 34, 57.

Kruschke, J. K. (2021). Bayesian analysis reporting guidelines. *Nature Human
Behaviour*, 5, 1282–1291.

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P.-C. (2021).
Rank-normalization, folding, and localization: An improved R-hat for
assessing convergence of MCMC. *Bayesian Analysis*, 16(2), 667–718.
