---
title: "Convergence diagnostics: what they mean and what failure looks like"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%")
options(mc.cores = 2)
```

```{r, message = FALSE}
library(bmbeR)
```

MCMC produces draws that are only guaranteed to represent the posterior in
the limit. Convergence diagnostics check whether *these* draws are good
enough. `check_convergence()` applies the current recommendations of the
Stan developers and Vehtari et al. (2021).

| Diagnostic | Default threshold | What failure means |
|---|---|---|
| Rank-normalised split-R-hat | ≤ 1.01 | Chains disagree with each other (or drift within themselves): they have not explored the same distribution. |
| Bulk-ESS | ≥ 100 × chains | Too few effectively independent draws to estimate means and medians reliably. |
| Tail-ESS | ≥ 100 × chains | Too few to estimate 5% and 95% quantiles, i.e. interval endpoints. |
| Divergent transitions | 0 | The sampler hit regions of high curvature it cannot explore; results may be biased (Betancourt, 2017). |
| E-BFMI | ≥ 0.3 per chain | The momentum resampling explores the energy distribution poorly (often heavy tails). |
| Max tree depth hits | reported only | An efficiency issue: the sampler needed very long trajectories. |

Two changes from older practice matter. The traditional R-hat threshold of
1.1 misses many failures that 1.01 catches, and the rank-normalised version
also detects chains that differ in scale or have heavy tails. Tail-ESS was
introduced because a good bulk-ESS can hide unreliable interval endpoints.

## A healthy fit

```{r}
data(kidiq, package = "rstanarm")
fit <- fit_model_with_prior(kidiq, kid_score ~ mom_iq + mom_hs, refresh = 0)
conv <- attr(fit, "bmb_convergence")
conv
head(conv$parameters)
```

Rank plots show, for each chain, a histogram of the ranks of its draws among
all chains. If the chains are sampling the same distribution, every
histogram is approximately uniform:

```{r}
plot(conv, type = "rank")
```

## When computation fails: often a modelling problem

The "folk theorem of statistical computing" (Gelman, 2008) says that when
you have computational problems, there is often a problem with your model.
A logistic regression on *separable* data, where a predictor perfectly
predicts the outcome, has no finite maximum-likelihood estimate. With a flat
prior the posterior is improper, and no sampler can succeed:

```{r}
set.seed(3)
sep <- data.frame(x = c(rnorm(20, -1), rnorm(20, 1)))
sep$y <- as.integer(sep$x > 0)  # x perfectly separates the classes

flat <- prior_config(intercept = prior_spec("flat"), slope = prior_spec("flat"))
fit_flat <- fit_model_with_prior(sep, y ~ x, family = binomial(),
                                 prior_config = flat, refresh = 0)
attr(fit_flat, "bmb_convergence")
```

R-hat, both ESS measures and the divergence check all fail, and almost every
iteration saturates the maximum tree depth as the sampler takes ever longer
trajectories. The trace plots show chains wandering off towards infinity,
each in its own direction:

```{r}
plot(attr(fit_flat, "bmb_convergence"), type = "trace")
```

Notice that `fit_model_with_prior()` warned but still returned the fit, so
it can be inspected. Pass `on_nonconvergence = "error"` to stop instead.

A weakly informative prior makes the posterior proper and the computation
easy. It also encodes something we genuinely believe: coefficients of a
logistic regression are not infinite.

```{r}
fit_wi <- fit_model_with_prior(sep, y ~ x, family = binomial(), refresh = 0)
attr(fit_wi, "bmb_convergence")
```

## Too few iterations

Even for a well-behaved model, too few iterations give unreliable estimates.
With 60 iterations per chain, warm-up is too short for adaptation:

```{r}
fit_short <- suppressWarnings(
  fit_model_with_prior(kidiq, kid_score ~ mom_iq + mom_hs, iter = 60,
                       refresh = 0, on_nonconvergence = "ignore")
)
attr(fit_short, "bmb_convergence")
```

The remedy is simply to run longer chains (the default `iter = 2000` is
ample for most regression models).

## Using the diagnostics in code

```{r}
conv <- check_convergence(fit, rhat_threshold = 1.01, min_ess = 400)
if (!isTRUE(conv$converged)) stop("Fix the computation before interpreting results.")
```

`check_convergence()` accepts both rstanarm (`stanreg`) and rstan (`stanfit`)
objects. `save_plots = TRUE` writes rank and trace plots to a PDF.
`generate_plot()` produces further diagnostic plots (histograms, densities,
autocorrelation) as a list of ggplot objects.

## References

Betancourt, M. (2017). A conceptual introduction to Hamiltonian Monte Carlo.
*arXiv:1701.02434*.

Gelman, A. (2008). The folk theorem of statistical computing. *Statistical
Modeling, Causal Inference, and Social Science* (blog), 13 May 2008.

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P.-C. (2021).
Rank-normalization, folding, and localization: An improved R-hat for
assessing convergence of MCMC. *Bayesian Analysis*, 16(2), 667–718.
