---
title: "Data-informed priors without double-dipping"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%")
options(mc.cores = 2)
```

```{r, message = FALSE}
library(bmbeR)
```

It is tempting to "let the data choose the prior": fit a quick regression,
centre each prior on the estimate, give it the estimate's standard error as
its scale, then fit the Bayesian model to the same data. bmbeR 1.x did
exactly this. This article shows why it fails and describes three published
alternatives implemented in `empirical_bayes_priors()`.

## The problem: counting the data twice

With a prior of \(N(\hat\beta, \mathrm{SE}^2)\) and a likelihood that is
approximately \(N(\hat\beta, \mathrm{SE}^2)\), the posterior is approximately
\(N(\hat\beta, \mathrm{SE}^2/2)\): the data have been used twice, so the
posterior claims twice the information the data contain. Credible intervals
are too narrow by a factor of \(\sqrt{2}\).

A quick simulation, using the normal approximation to the posterior, shows the
consequence for the coverage of nominal 95% intervals:

```{r}
set.seed(1)
n <- 50
sims <- replicate(4000, {
  x <- rnorm(n)
  y <- 1 + 0.5 * x + rnorm(n)
  est <- summary(lm(y ~ x))$coefficients["x", ]
  b <- est[[1]]; se <- est[[2]]
  post_sd <- c(
    flat             = se,                                # likelihood only
    same_data_prior  = se / sqrt(2),                      # N(b, se^2) prior, bmbeR 1.x
    unit_information = se * sqrt(n / (n + 1))             # N(b, n se^2) prior
  )
  abs(b - 0.5) <= 1.96 * post_sd
})
coverage <- rowMeans(sims)
round(coverage, 3)
```

Centring a tight prior on the same data drops the coverage of "95%" intervals
to `r round(100 * coverage[["same_data_prior"]])`%, while the unit-information
prior, described next, is as well calibrated as the likelihood alone.

## Unit-information priors

The unit-information prior (Kass & Wasserman, 1995) is centred on the
estimate but carries the information of a *single* observation: its variance
is \(n\) times the sampling variance. It regularises extreme estimates and
makes the prior's scale automatically appropriate to the units of the data,
while adding only \(1/n\) of the data's information. It is the prior
implicitly behind BIC.

```{r}
data(kidiq, package = "rstanarm")
ui <- empirical_bayes_priors(kidiq, kid_score ~ mom_iq + mom_hs)
ui
```

The output plugs straight into `fit_model_with_prior()`. The intercept prior
is placed on the intercept at the predictor means, which is the parameter
rstanarm puts its intercept prior on.

```{r}
fit_ui <- fit_model_with_prior(kidiq, kid_score ~ mom_iq + mom_hs,
                               prior_config = ui, refresh = 0)
prior_sensitivity(fit_ui)$summary[, c("variable", "contraction", "diagnosis")]
```

The posterior contraction is close to \(n/(n+1) =\)
`r round(nrow(kidiq) / (nrow(kidiq) + 1), 4)`, as theory predicts: the prior
contributes about one observation's worth of information.

For logistic, Poisson and other GLMs, `empirical_bayes_priors()` fits a GLM
in the same family, so the priors are on the correct (link) scale. Always
pass the same `formula` and `family` you will fit; `fit_model_with_prior()`
checks that the coefficients match.

## Empirical Bayes shrinkage

When a model has many coefficients of comparable size, a classic result
(James & Stein, 1961; Efron & Morris, 1973) is that shrinking all estimates
towards zero by a common, *data-estimated* amount beats estimating each on
its own. `method = "eb_shrinkage"` implements the parametric empirical Bayes
version (Morris, 1983):

1. put coefficients on a common scale by multiplying by the predictors'
   standard deviations;
2. estimate the spread of effects \(\tau\) by maximising the marginal
   likelihood of \(b_j \sim N(0, \tau^2 + \mathrm{SE}_j^2)\);
3. give each coefficient a zero-centred prior with scale \(\tau / \mathrm{sd}(x_j)\).

To see the benefit we simulate data with 30 weak predictors and only 80
observations, then score predictions on a large test set:

```{r}
set.seed(2024)
k <- 30
beta <- rnorm(k, 0, 0.3)
simulate <- function(n) {
  X <- matrix(rnorm(n * k), n, k, dimnames = list(NULL, sprintf("x%02d", 1:k)))
  data.frame(X, y = 1 + drop(X %*% beta) + rnorm(n, 0, 2))
}
train <- simulate(80)
test  <- simulate(2000)
f <- reformulate(sprintf("x%02d", 1:k), "y")

eb <- empirical_bayes_priors(train, f, method = "eb_shrinkage")
attr(eb, "tau")  # estimated spread of the (standardised) effects; truth is 0.3

fits <- list(
  rstanarm_default = fit_model_with_prior(train, f, refresh = 0),
  unit_information = fit_model_with_prior(train, f, prior_config = empirical_bayes_priors(train, f),
                                          refresh = 0),
  eb_shrinkage     = fit_model_with_prior(train, f, prior_config = eb, refresh = 0)
)
perf <- lapply(fits, evaluate_model_performance, data_test = test)
compare_performance(perf)
sapply(fits, function(fit) sqrt(mean((coef(fit)[-1] - beta)^2)))  # coefficient RMSE
```

Shrinkage improves both the held-out log score and the accuracy of the
coefficients. It needs at least three coefficients and works best when they
are exchangeable (similar in kind). When the data contain little information
about the spread of effects, the marginal-likelihood estimate of \(\tau\)
can collapse to zero; bmbeR then floors it and warns:

```{r}
eb_mtcars <- empirical_bayes_priors(
  mtcars, mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb,
  method = "eb_shrinkage"
)
```

Empirical Bayes still uses the data twice (to estimate \(\tau\) and to fit),
but only through one hyperparameter; the resulting understatement of
uncertainty is small when there are many coefficients.

## Power priors from historical data

When a previous study measured the same effects, its data can inform the
prior legitimately because they are *different* data. The power prior
(Ibrahim & Chen, 2000) raises the historical likelihood to a power
\(a_0 \in (0, 1]\) that discounts it: \(a_0 = 1\) pools the studies,
\(a_0 = 0.5\) counts each historical observation as half an observation.
bmbeR uses the normal approximation: the prior is centred on the historical
estimates with variances inflated by \(1/a_0\).

Here we split `kidiq` into a "historical" and a "current" study:

```{r}
set.seed(7)
hist_rows <- sample(nrow(kidiq), 234)
historical <- kidiq[hist_rows, ]
current    <- kidiq[-hist_rows, ]

fits_pp <- list(no_borrowing = fit_model_with_prior(current, kid_score ~ mom_iq + mom_hs,
                                                    refresh = 0))
for (a0 in c(0.25, 0.5, 1)) {
  pp <- suppressMessages(empirical_bayes_priors(historical, kid_score ~ mom_iq + mom_hs,
                                                method = "power", a0 = a0))
  fits_pp[[paste0("a0 = ", a0)]] <- fit_model_with_prior(current, kid_score ~ mom_iq + mom_hs,
                                                         prior_config = pp, refresh = 0)
}
t(sapply(fits_pp, function(fit) {
  d <- as.matrix(fit)
  c(mom_iq_mean = mean(d[, "mom_iq"]), mom_iq_sd = sd(d[, "mom_iq"]),
    mom_hs_mean = mean(d[, "mom_hs"]), mom_hs_sd = sd(d[, "mom_hs"]))
}))
```

Borrowing more (larger \(a_0\)) narrows the posterior. Choosing \(a_0\) is a
substantive judgement about how similar the studies are; run
`prior_sensitivity()` on the result, since a historical study that differs
from the current one shows up as a prior-data conflict:

```{r}
prior_sensitivity(fits_pp[["a0 = 1"]])$summary[, c("variable", "contraction", "conflict_z", "diagnosis")]
```

Never use the analysed data as "historical" data: that is the double-dipping
of the first section, with \(a_0 = 1\).

## Which method when?

| Situation | Method |
|---|---|
| No external information; want a scale-aware, minimally informative prior | `"unit_information"` |
| Many exchangeable coefficients, few observations each | `"eb_shrinkage"` |
| A previous study measured the same effects | `"power"` with historical data |
| Genuine substantive knowledge | `prior_spec()` directly |

Add `type = "student_t"` to any method to give the priors heavier tails, so
that the data can override them if prior and data turn out to conflict
(O'Hagan & Pericchi, 2012).

## References

Efron, B., & Morris, C. (1973). Stein's estimation rule and its
competitors—an empirical Bayes approach. *JASA*, 68(341), 117–130.

Ibrahim, J. G., & Chen, M.-H. (2000). Power prior distributions for
regression models. *Statistical Science*, 15(1), 46–60.

James, W., & Stein, C. (1961). Estimation with quadratic loss. *Proceedings
of the Fourth Berkeley Symposium*, 1, 361–379.

Kass, R. E., & Wasserman, L. (1995). A reference Bayesian test for nested
hypotheses and its relationship to the Schwarz criterion. *JASA*, 90(431),
928–934.

Morris, C. N. (1983). Parametric empirical Bayes inference: Theory and
applications. *JASA*, 78(381), 47–55.

O'Hagan, A., & Pericchi, L. (2012). Bayesian heavy-tailed models and conflict
resolution: A review. *Brazilian Journal of Probability and Statistics*,
26(4), 372–401.
