---
title: "Predictive evaluation with proper scoring rules"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%")
options(mc.cores = 2)
```

```{r, message = FALSE}
library(bmbeR)
library(ggplot2)
```

A Bayesian model predicts a *distribution*, not a number. Scoring only the
posterior mean (RMSE, accuracy) ignores whether the model's uncertainty is
honest. **Proper scoring rules** reward a forecast for being both accurate and
calibrated: they are optimised in expectation only by reporting your true
predictive distribution (Gneiting & Raftery, 2007).
`evaluate_model_performance()` reports:

| Metric | Outcome | Better | Notes |
|---|---|---|---|
| `ELPD` | all | higher | log predictive density on the test set; the out-of-sample analogue of `elpd_loo` |
| `CRPS` | continuous / count | lower | in the units of the outcome; reduces to absolute error for a point forecast |
| `Brier` | binary | lower | mean squared error of predicted probabilities |
| `Coverage` | continuous / count | ≈ `prob` | calibration of the predictive intervals |
| `AUC` | binary | higher | discrimination only (not a proper score) |
| `RMSE`, `MAE`, `Accuracy`, `Precision`, `Recall`, `F1_Score` | | | familiar point-prediction metrics |

## Binary outcomes: the wells data

```{r}
data(wells, package = "rstanarm")
wells$dist100 <- wells$dist / 100
set.seed(123)
test_rows <- sample(nrow(wells), 1000)
train <- wells[-test_rows, ]
test  <- wells[test_rows, ]

fit_dist <- fit_model_with_prior(train, switch ~ dist100, family = binomial(),
                                 refresh = 0)
fit_full <- fit_model_with_prior(train, switch ~ dist100 + arsenic + educ,
                                 family = binomial(), refresh = 0)
perf_dist <- evaluate_model_performance(fit_dist, test)
perf_full <- evaluate_model_performance(fit_full, test)
perf_full
```

Classification metrics are computed from the posterior mean *probability*
of switching, thresholded at 0.5 (see `threshold`). Accuracy alone can hide a
lot: the Brier score and ELPD also penalise overconfident probabilities.

### Comparing models on the same test set

Because both models are scored on the same households, the difference in
ELPD should be assessed with the standard error of the *paired*
(per-observation) differences, not by comparing the two ELPDs and their
separate standard errors:

```{r}
compare_performance(distance_only = perf_dist, full = perf_full)
```

A difference of several standard errors is a clear improvement.

### Calibration

For a calibrated model, among households given a predicted probability of
about 0.7, about 70% should switch. A reliability plot checks this:

```{r}
p <- colMeans(rstanarm::posterior_epred(fit_full, newdata = test))
bins <- cut(p, breaks = quantile(p, 0:10 / 10), include.lowest = TRUE)
calib <- data.frame(predicted = tapply(p, bins, mean),
                    observed  = tapply(test$switch, bins, mean))
ggplot(calib, aes(predicted, observed)) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.5, colour = "#1b6ca8") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Mean predicted probability (decile bins)", y = "Observed share switching",
       title = "Reliability plot") +
  theme_bw()
```

## Continuous outcomes

For continuous outcomes the CRPS and interval coverage complement RMSE:

```{r}
data(kidiq, package = "rstanarm")
set.seed(1)
test_rows <- sample(nrow(kidiq), 100)
fit_kid <- fit_model_with_prior(kidiq[-test_rows, ], kid_score ~ mom_iq * mom_hs,
                                refresh = 0)
evaluate_model_performance(fit_kid, kidiq[test_rows, ], prob = 0.9)
```

Coverage close to the nominal 90% means the predictive intervals are about
the right width; much lower coverage signals overconfidence.

## Held-out data or LOO?

When no test set is available, PSIS-LOO cross-validation (Vehtari et al.,
2017) estimates the same ELPD from the training data alone:

```{r}
loo::loo(fit_full)
```

`workflow_report()` reports LOO automatically and includes held-out metrics
when you pass them via `performance =`.

## References

Brier, G. W. (1950). Verification of forecasts expressed in terms of
probability. *Monthly Weather Review*, 78(1), 1–3.

Gneiting, T., & Raftery, A. E. (2007). Strictly proper scoring rules,
prediction, and estimation. *JASA*, 102(477), 359–378.

Vehtari, A., Gelman, A., & Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5), 1413–1432.
