---
title: "Prior sensitivity: power-scaling diagnostics and refitting"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%")
options(mc.cores = 2)
```

```{r, message = FALSE}
library(bmbeR)
data(kidiq, package = "rstanarm")
```

Every Bayesian analysis should ask *how much did the prior matter?* bmbeR
offers two complementary answers:

* `prior_sensitivity()`: **local** diagnostics computed from a single fit, with
  no refitting, by power-scaling the prior and the likelihood (Kallioinen et
  al., 2023).
* `sensitivity_analysis()`: a **global** check that refits the model under
  alternative priors and compares posteriors and predictive performance.

## Power-scaling

Power-scaling raises the prior (or the likelihood) to a power \(\alpha\):

$$p_\alpha(\theta \mid y) \propto p(\theta)^\alpha \, p(y \mid \theta).$$

\(\alpha > 1\) strengthens the prior, \(\alpha < 1\) weakens it. If the
posterior barely changes as \(\alpha\) moves away from 1, the prior is not
driving the results.

### Local sensitivity is a posterior covariance

Differentiating a posterior expectation with respect to \(\log\alpha\) at
\(\alpha = 1\) gives a posterior covariance (Giordano et al., 2018):

$$\left.\frac{d\,E_\alpha[f(\theta)]}{d\log\alpha}\right|_{\alpha=1}
= \mathrm{Cov}\big(f(\theta),\, \log p(\theta)\big).$$

So the sensitivities can be computed exactly (up to Monte Carlo error) from
the posterior draws you already have. bmbeR reports, for each parameter:

* `prior_mean_sens` \(= \mathrm{Cov}(\theta, \log p(\theta))/\sigma\): the
  change in the posterior mean, in posterior SDs, per unit of \(\log\alpha\);
* `prior_sd_sens` \(= \mathrm{Cov}((\theta-\mu)^2, \log p(\theta))/(2\sigma^2)\):
  the relative change in the posterior SD;
* `lik_mean_sens`, `lik_sd_sens`: the same with the log-likelihood.

### What the numbers mean

In a normal model with prior \(N(m_0, s_0^2)\) and likelihood
\(N(\bar y, s^2)\), let \(w = \sigma^2/s_0^2\) be the share of posterior
precision contributed by the prior. Then

| Quantity | Normal-model value | Interpretation |
|---|---|---|
| `prior_sd_sens` | \(-w/2\) | prior's share of the posterior precision (halved) |
| `lik_sd_sens` | \(-(1-w)/2\) | data's share of the posterior precision (halved) |
| `prior_mean_sens` | \(w(1-w)(m_0-\bar y)/\sigma\) | how hard the prior pulls the mean |
| `lik_mean_sens` | \(-\)`prior_mean_sens` | mirror image; no separate information |
| `conflict_z` | \((m_0-\bar y)/\sqrt{s_0^2+s^2}\) | prior-data conflict (Box, 1980) |
| `contraction` | \(1 - \sigma^2/\mathrm{Var}_{prior}\) | how much the data updated the prior (Schad et al., 2021) |

`conflict_z` is computed as
`prior_mean_sens / (2 * sqrt(prior_sd_sens * lik_sd_sens))`, which is exactly
the classic prior predictive z-score in the normal case. That separates a
prior that is *informative but compatible* with the data from one that
*conflicts* with them, which a single sensitivity number cannot do.

### Diagnoses

| Diagnosis | Condition | What to do |
|---|---|---|
| `-` | prior influence < `threshold` | Nothing: the data dominate. |
| `weak likelihood` | prior influential, `|lik_sd_sens|` < `threshold` | The data say little about this parameter; the prior must be defensible, or collect more data. |
| `prior-data conflict` | prior influential, likelihood informative, `|conflict_z|` ≥ 2 | Prior and data disagree: revisit the prior (or the model), or use heavier tails. |
| `informative prior` | prior influential, likelihood informative, `|conflict_z|` < 2 | Fine if the prior is justified; report it. |

The default `threshold = 0.05` corresponds to the prior contributing about 10%
of the posterior precision in the normal model.

## Four scenarios

We fit `kid_score ~ mom_iq` under four priors for the slope and examine only
the slope's prior (`scale_priors = "mom_iq"`):

```{r}
set.seed(5)
few <- kidiq[sample(nrow(kidiq), 8), ]
scenarios <- list(
  "weakly informative (default)" = list(kidiq, NULL),
  "informative, agrees"          = list(kidiq, prior_config(slope = prior_spec("normal", 0.6, 0.06))),
  "informative, conflicts"       = list(kidiq, prior_config(slope = prior_spec("normal", 1.2, 0.05))),
  "informative, 8 observations"  = list(few,   prior_config(slope = prior_spec("normal", 0.6, 0.05)))
)
fits <- lapply(scenarios, function(s)
  fit_model_with_prior(s[[1]], kid_score ~ mom_iq, prior_config = s[[2]], refresh = 0))
tab <- do.call(rbind, lapply(names(fits), function(nm) {
  s <- prior_sensitivity(fits[[nm]], pars = "mom_iq", scale_priors = "mom_iq")$summary
  data.frame(scenario = nm, s[, -1])
}))
rownames(tab) <- NULL
num <- vapply(tab, is.numeric, logical(1))
tab[num] <- lapply(tab[num], round, 3)
tab
```

* With the default prior the data dominate.
* The agreeing informative prior roughly halves the posterior variance
  (contraction about 0.5) without moving the mean: `informative prior`.
* The conflicting prior drags the slope far from the data's estimate of about
  0.6, and `conflict_z` is far beyond 2.
* With eight observations the data contribute very little precision, so the
  posterior is essentially the prior: `weak likelihood`.

The prior-vs-posterior plot of the conflicting case shows the posterior
wedged between the prior and the likelihood:

```{r}
plot_prior_posterior(fits[["informative, conflicts"]], pars = "mom_iq")
```

## Finite perturbations and their reliability

The derivative describes infinitesimal changes. `prior_sensitivity()` also
estimates the posterior under finite powers (by default \(\alpha = 0.8\) and
\(1.25\)) by Pareto-smoothed importance sampling (Vehtari et al., 2024), and
flags estimates whose Pareto \(\hat k\) is too high to trust:

```{r}
ps <- prior_sensitivity(fits[["informative, conflicts"]])
ps
plot(ps)
```

Weakening (\(\alpha < 1\)) is the harder direction for importance sampling.
For large perturbations, refit with `sensitivity_analysis()` instead.

## Joint versus selective power-scaling

By default the *joint* prior of all parameters is power-scaled, so a
conflict in one prior can surface in other parameters. In the output above,
`sigma` is flagged as well as `mom_iq`: the misplaced slope leaves systematic
misfit that a larger residual SD absorbs. To locate the source, scale one
prior at a time:

```{r}
for (p in c("(Intercept)", "mom_iq", "sigma")) {
  d <- prior_sensitivity(fits[["informative, conflicts"]], scale_priors = p)$summary
  cat(sprintf("scaling the %-12s prior -> %s\n", p,
              paste(sprintf("%s: %s", d$variable, d$diagnosis), collapse = "; ")))
}
```

Only scaling the `mom_iq` prior produces the conflict.

## Refitting under alternative priors

Local diagnostics describe the posterior you obtained. A global check refits
the model under a set of plausible alternative priors:

```{r}
sens <- sensitivity_analysis(
  kidiq, kid_score ~ mom_iq + mom_hs,
  prior_configurations = list(
    default          = NULL,
    unit_information = empirical_bayes_priors(kidiq, kid_score ~ mom_iq + mom_hs),
    sceptical        = prior_config(slope = prior_spec("normal", 0, c(0.1, 1))),
    heavy_tailed     = prior_config(slope = prior_spec("student_t", 0, c(1, 10), df = 3))
  ),
  refresh = 0
)
sens
plot(sens, pars = c("mom_iq", "mom_hs"))
```

```{r, include = FALSE}
sc <- sens$summary[sens$summary$config == "sceptical" & sens$summary$variable %in% c("mom_iq", "mom_hs"), ]
cmp <- sens$comparison["sceptical", ]
```

The table reports each configuration's posterior means with the shift from
the reference configuration in reference posterior SDs (shifts of at least
`shift_threshold = 0.5` are marked `!`). The LOO comparison shows whether the
priors also change out-of-sample predictive performance; differences smaller
than about two standard errors are not meaningful (Vehtari et al., 2017).

Read both parts. Here the sceptical prior moves the two slopes by
`r paste(sprintf("%.1f", abs(sc$shift_sd)), collapse = " and ")` posterior
SDs, a material change in the estimates, while its LOO difference of
`r sprintf("%.1f", cmp[["elpd_diff"]])` (SE `r sprintf("%.1f", cmp[["se_diff"]])`)
is `r if (abs(cmp[["elpd_diff"]]) < 2 * cmp[["se_diff"]]) "too small to distinguish it from the others" else "a clear loss of predictive performance"`.
A prior can change an estimate substantially while barely changing
predictions.

## Relation to priorsense

The [priorsense](https://github.com/n-kall/priorsense) package implements
power-scaling sensitivity analysis (Kallioinen et al., 2023) for brms,
cmdstanr, rstan, JAGS and NIMBLE models, and for draws with user-supplied
log-prior and log-likelihood values. It measures sensitivity with the
cumulative Jensen–Shannon distance. rstanarm fits do not store log-prior
evaluations, so bmbeR reconstructs them from `rstanarm::prior_summary()`
(including autoscaling and the centred intercept). It reports the mean and
SD sensitivities, whose normal-model interpretations are given above, and
the `conflict_z` statistic. The two approaches answer the same question and
should agree on which parameters are sensitive, though their numerical
scales differ.

## References

Box, G. E. P. (1980). Sampling and Bayes' inference in scientific modelling
and robustness. *JRSS A*, 143(4), 383–430.

Giordano, R., Broderick, T., & Jordan, M. I. (2018). Covariances,
robustness, and variational Bayes. *JMLR*, 19(51), 1–49.

Kallioinen, N., Paananen, T., Bürkner, P.-C., & Vehtari, A. (2023). Detecting
and diagnosing prior and likelihood sensitivity with power-scaling.
*Statistics and Computing*, 34, 57.

Schad, D. J., Betancourt, M., & Vasishth, S. (2021). Toward a principled
Bayesian workflow in cognitive science. *Psychological Methods*, 26(1),
103–126.

Vehtari, A., Gelman, A., & Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5), 1413–1432.

Vehtari, A., Simpson, D., Gelman, A., Yao, Y., & Gabry, J. (2024). Pareto
smoothed importance sampling. *JMLR*, 25(72), 1–58.
