---
title: "Choosing priors: scale, autoscaling and prior predictive checks"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96, out.width = "100%")
options(mc.cores = 2)
```

```{r, message = FALSE}
library(bmbeR)
```

A prior is a statement about plausible parameter values, and a regression
coefficient's plausible values depend entirely on the units of the outcome
and the predictors. This article shows what goes wrong when that is ignored,
the options for getting it right, and how prior predictive checks catch
problems before any model is fitted.

## Why scale matters

`mtcars` records fuel efficiency (`mpg`) and weight (`wt`, in 1000 lbs). We
fit the same model twice: once with `mpg`, and once with the outcome
expressed in hundredths of a mile per gallon (`mpg * 100`). Nothing about the
data has changed except the units, so the second slope should be exactly 100
times the first.

First with a *fixed-scale* prior, `student_t(3, 0, 2.5)` on every
coefficient, which was the default in bmbeR 1.x:

```{r}
d <- mtcars
d$mpg100 <- d$mpg * 100
fixed <- prior_config(intercept = prior_spec("student_t", 0, 2.5, df = 3),
                      slope     = prior_spec("student_t", 0, 2.5, df = 3))

fit_mpg    <- fit_model_with_prior(d, mpg ~ wt, prior_config = fixed,
                                   refresh = 0, on_nonconvergence = "ignore")
fit_mpg100 <- fit_model_with_prior(d, mpg100 ~ wt, prior_config = fixed,
                                   refresh = 0, on_nonconvergence = "ignore")
rbind(
  "OLS, mpg"            = coef(lm(mpg ~ wt, d)),
  "fixed prior, mpg"    = coef(fit_mpg),
  "OLS, mpg*100"        = coef(lm(mpg100 ~ wt, d)),
  "fixed prior, mpg*100" = coef(fit_mpg100)
)
```

On the original scale the prior is harmless. On the rescaled outcome the
same prior says "the slope is almost certainly within a few units of zero",
while the data say it is around −530. The prior wins: the model attributes
everything to noise (look at `sigma`).

```{r}
plot_prior_posterior(fit_mpg100)
```

The MCMC diagnostics do not flag this, because the chains mix perfectly well
around the wrong answer:

```{r}
attr(fit_mpg100, "bmb_convergence")$converged
```

Two checks do catch it. **Before fitting**, a prior predictive check shows
that these priors expect outcomes centred on zero, far from any plausible
value of `mpg * 100`:

```{r}
prior_predictive_check(d, mpg100 ~ wt, prior_config = fixed,
                       plausible_range = c(500, 5000), refresh = 0)
```

**After fitting**, `prior_sensitivity()` shows that the posterior is
dominated by the prior. The posterior has found a self-consistent but wrong
explanation: with a huge residual SD the data carry almost no information
about the slope, so the local diagnosis is "weak likelihood", and the slope's
posterior contraction is close to zero despite 32 observations with a strong
relationship.

```{r}
s <- prior_sensitivity(fit_mpg100)$summary
s[, c("variable", "mean", "contraction", "diagnosis")]
```

Power-scaling diagnostics are *local*: they describe the posterior you
obtained, not the one you should have obtained. That is why the prior
predictive check, which is a global check, comes first.

## Four ways to get the scale right

1. **Standardise** outcome and predictors, then use unit-scale priors such as
   `normal(0, 1)`. Simple and transparent, but coefficients must be
   back-transformed for interpretation.
2. **Autoscale.** `prior_spec(..., autoscale = TRUE)` lets rstanarm rescale
   the prior by the standard deviations of the outcome and predictors
   (Gelman et al., 2008). Components you leave `NULL` in `prior_config()` use
   rstanarm's autoscaled defaults.
3. **Use substantive knowledge** in the natural units, as in the
   *Get started* vignette.
4. **Derive priors from data** defensibly with `empirical_bayes_priors()`
   (see the *Data-informed priors* article).

With the (autoscaled) defaults, the two fits agree up to the change of units:

```{r}
fit_default <- fit_model_with_prior(d, mpg100 ~ wt, refresh = 0)
coef(fit_default)["wt"] / coef(fit_mpg)["wt"]
```

`rstanarm::prior_summary()` shows what autoscaling did:

```{r}
rstanarm::prior_summary(fit_default)
```

## Prior predictive checks

A prior predictive check simulates outcomes from the priors alone (the
likelihood is switched off) and asks whether they look like data that could
plausibly have been observed (Gabry et al., 2019). It is the most direct way
to see what a set of priors *means*.

### A continuous outcome

```{r}
data(kidiq, package = "rstanarm")
pc_default <- prior_predictive_check(kidiq, kid_score ~ mom_iq + mom_hs,
                                     plausible_range = c(0, 200), refresh = 0)
pc_default
```

rstanarm's defaults are deliberately weak: a noticeable share of simulated
test scores are negative or above 200. That is often acceptable (weak priors
let the data speak), but the check makes the trade-off visible. Priors based
on what test scores can look like (see the *Get started* vignette) keep the
simulations in a sensible range.

```{r}
plot(pc_default)
```

### Logistic regression: "non-informative" priors are informative

For binary outcomes, wide priors on the logit scale are a classic trap. The
`wells` data (Gelman & Hill, 2007) record whether households in Bangladesh
switched away from an unsafe well. Consider three sets of priors for a model
with three predictors:

```{r}
data(wells, package = "rstanarm")
wells$dist100 <- wells$dist / 100
f <- switch ~ dist100 + arsenic + educ

wide <- prior_config(intercept = prior_spec("normal", 0, 10),
                     slope     = prior_spec("normal", 0, 10))
narrow <- prior_config(slope = prior_spec("normal", 0, 0.5, autoscale = TRUE))

checks <- list(
  "normal(0, 10)"              = prior_predictive_check(wells, f, binomial(), wide, refresh = 0),
  "rstanarm default"           = prior_predictive_check(wells, f, binomial(), NULL, refresh = 0),
  "normal(0, 0.5), autoscaled" = prior_predictive_check(wells, f, binomial(), narrow, refresh = 0)
)
sapply(checks, function(x) x$summary$share_extreme)
```

The numbers are the share of prior predictive success probabilities below
5% or above 95%. The "vague" `normal(0, 10)` prior claims that almost every
household's decision is a foregone conclusion. Even the weakly informative
default does this for a large share of households, because the variances of
independent coefficient priors add up on the logit scale as predictors
accumulate (Gelman et al., 2020). Tighter, autoscaled priors spread the
probabilities more evenly, which matches what we actually believe before
seeing the data.

```{r}
plot(checks[["normal(0, 10)"]])
```

## Checklist

* Write priors in the units of your data, or standardise, or autoscale.
* Run `prior_predictive_check()` with a `plausible_range` you can defend.
* After fitting, run `prior_sensitivity()`: a `prior-data conflict` or a
  small `contraction` tells you the prior is doing more work than you may
  have intended.

## References

Gabry, J., Simpson, D., Vehtari, A., Betancourt, M., & Gelman, A. (2019).
Visualization in Bayesian workflow. *JRSS A*, 182(2), 389–402.

Gelman, A., & Hill, J. (2007). *Data Analysis Using Regression and
Multilevel/Hierarchical Models*. Cambridge University Press.

Gelman, A., Jakulin, A., Pittau, M. G., & Su, Y.-S. (2008). A weakly
informative default prior distribution for logistic and other regression
models. *The Annals of Applied Statistics*, 2(4), 1360–1383.

Gelman, A., Vehtari, A., Simpson, D., et al. (2020). Bayesian workflow.
*arXiv:2011.01808*.
